\huge The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-06-23 Writeup/kohler_clarke_2021.tex
minor revisions, no longer tracked changes

\begin{abstract}
\textcolor{ForestGreen}{Symmetries are present at many scales in natural scenes. Humans and other animals are highly sensitive to visual symmetry, and symmetry contributes to numerous domains of visual perception. The four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined into exactly 17 distinct regular textures. These \textit{wallpaper groups} represent the complete set of symmetries in 2D images. The current study seeks to provide a more comprehensive description of responses to symmetry in the human visual system, by collecting both brain imaging (Steady-State Visual Evoked Potentials measured using high-density EEG) and behavioral (symmetry detection thresholds) data using the entire set of wallpaper groups. This allows us to probe the hierarchy of complexity among wallpaper groups, in which simpler groups are subgroups of more complex ones. We find that both behavior and brain activity preserve the hierarchy almost perfectly: Subgroups consistently produce lower amplitude symmetry-specific responses in visual cortex and require longer presentation durations to be reliably detected. These findings expand our understanding of symmetry perception by showing that the human brain encodes symmetries with a high level of precision and detail. This opens new avenues for research on how fine-grained representations of regular textures contribute to natural vision.}
\end{abstract}

Symmetries are abundant in natural and man-made environments, due to a complex interplay of physical forces that govern pattern formation in nature. \textcolor{ForestGreen}{Sensitivity to symmetry has been demonstrated in a number of species, includes bees \citep{giurfa_insect_symmetry_1996}, fish \citep{morris_swordtail_symmetry_1998,schluter_sailfin_symmetry_1998}, birds \citep{RN1337,swaddle_symmetry_finches_1994} and dolphins \citep{von_fersen_dolphin_symmetry_1992}, and may be used as a cue for mate selection in many species \citep{swaddle_asymmetry_review_1999} including humans \citep{rhodes_facial_1998}. Humans cultures have created and appreciated s}ymmetrical patterns throughout history, and since the gestalt movement of the early 20th century, symmetry has been recognized as important for visual perception. Symmetry contributes to the perception of shapes \citep{RN1311,RN1682}, scenes \citep{apthorp_symmetry_numerosity} and surface properties \citep{RN1166}. \textcolor{ForestGreen}{This literature is almost exclusively based on stimuli in which one or more symmetry axes are placed at a single point in the image. Focus has been on} mirror symmetry or \textit{reflection}, \textcolor{ForestGreen}{with relatively few studies} including the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection} \citep{wagemans_1998} \textcolor{ForestGreen}{- perhaps because reflection has been found to be more perceptually salient \citep{mach_1959,royer_1981,palmer_1991,ogden_2016,hamada_1988} and produce more brain activity \citep{makin_2013,makin_2014,makin_2012,wright_2015}. In the current study, we take a different approach by investigating visual processing of regular textures in which combinations of the four fundamental symmetries tile the 2D plane.}

In the two spatial dimensions relevant for images, symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \citep{RN1562,RN1563,RN1425}. Previous work on a subset of four of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \citep{RN1725}. This effect was also robust \textcolor{ForestGreen}{when symmetry responses were measured with electroencephalography (EEG) using both Steady-State Visual Evoked Potentials (SSVEPs)\citep{RN1725} and Event-Related Potentials \citep{RN1959}. The SSVEP technique uses periodic visual stimulation to produce a periodic brain response that is confined to integer multiples of the stimulation frequency known as harmonics. SSVEP response harmonics can be isolated in the frequency domain and depending on the specific design, different harmonics will express different aspects of the brain response. \citep{RN1540}.} Here we extend on the previous work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \citep{RN1425}. The wallpaper groups differ in the number and kind of these transformations \textcolor{ForestGreen}{and we can uniquely refer to different groups using crystallographic notation. In brief, most groups are notated by $PXZ$, where $X\in\{1,2,3,4,6\}$ indicates the highest order of rotation symmetry and $Z\in\{m, g\}$ indicates whether the pattern contains reflection (m) or glide reflection (g). For example, $P4$ contains rotation of order 4, while $P4MM$ contains rotation of order 4 and two reflections. By convention, many of the groups are given shortened names: for example, $P4MM$ is usually referred to as $P4M$, as the second reflection can be deduced from the presence of rotation of order 4 alongside a reflection. Two of the groups start with a $C$ rather than a $P$, ($CM$ and $CMM$) which indicates that the symmetries are specified relative to a cell that itself contains repetition. Full details of the naming convention can be found on \href{https://en.wikipedia.org/wiki/Wallpaper_group}{wikipedia} and examples of the wallpaper groups are shown in Figures \ref{fig:subgroups} and \ref{fig:summary}.}

In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \citep{RN1425}. The full list of subgroup relationships is listed in Section 1.4.2 of the Supplementary Material. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the supergroup \citep{RN1425}. As an example, let us consider groups \textit{P2} and \textit{P6} \textcolor{ForestGreen}{(see Figure \ref{fig:subgroups}B)}. If we ignore the translations in two directions that both groups share, group \textit{P6} consists of the set of rotations \{$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$\}, in which \textit{P2} \{$0^{\circ}$, $180^{\circ}$\} is contained. \textit{P2} is thus a subgroup of \textit{P6}, and \textit{P6} can be generated by combining \textit{P2} with rotations \{$0^{\circ}$, $120^{\circ}$, $240^{\circ}$\}. Because \textit{P2} is repeated three times in \textit{P6}, \textit{P2} is a subgroup of \textit{P6} with index 3 \citep{RN1425}. \textcolor{ForestGreen}{Similarly, \textit{PMM} contains two reflections and rotations \{$0^{\circ}$, $180^{\circ}$\}. \textit{PMM} can be generated by adding an additional reflection to both \textit{P2} (\{$0^{\circ}$, $180^{\circ}$\}) and \textit{PM} (one reflection), so \textit{P2} and \textit{PM} are both subgroups of \textit{PMM} with index 2 (see Figure \ref{fig:subgroups}C).} The 17 wallpaper groups thus obey a hierarchy of complexity where simpler groups are subgroups of more complex ones \citep{RN1711}.

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure1.pdf}
\caption{\textcolor{ForestGreen}{Subgroup relationships with indices 2 (solid lines) and 3 (dashed line) are shown in (A). All other relationships can be inferred by identifying the shortest path through the hierarchy, and multiplying the subgroup indices. For example, \textit{P1} is related to \textit{P6} through \textit{P6}\textrightarrow\textit{P3} (index 2) and \textit{P3}\textrightarrow\textit{P1} (index 3) so \textit{P1} is also a subgroup of \textit{P6} with index $3 \times 2 = 6$. We also show enlarged versions of some of the subgroup relationships involving \textit{P6} (B, shown in red) and \textit{PMM} (C, shown in blue) and highlight the symmetries within the subgroups to emphasize how the supergroup can be generated by adding additional transformations to the subgroup.}}
\label{fig:subgroups}
\end{figure}

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \citep{Sasaki_2005, Tyler_2005} and rotation \citep{RN1725}, the visual system represents the hierarchical structure of the 17 wallpaper groups, and thus every \textcolor{ForestGreen}{combination} of the four fundamental symmetries (rotation, reflection, translation, glide reflection) which comprise the set of regular textures. Both SSVEP amplitudes and symmetry detection thresholds preserve the hierarchy of complexity among the wallpaper groups that is captured by the subgroup relationships \citep{RN1711}. For the SSVEP, this remarkable consistency was specific to the odd harmonics of the stimulus frequency that are known to capture the symmetry-specific response \citep{RN1725} and to electrodes in a region-of-interest (ROI) over occipital cortex. When the same analysis was done using the odd harmonics from electrodes over parietal cortex (Supplementary Figure 3.3) or even harmonics from electrodes over occipital cortex (Supplementary Figure 3.4), the data was substantially less consistent with the subgroup relationships (yellow and green lines, Figure 5).

\textcolor{ForestGreen}{The current study uses 16 distinct wallpaper groups, while previous neuroimaging studies focused on a subset of 4 \citep{RN1725,RN1959}. This represents a significant conceptual advance, because it makes it possible to investigate the complete subgroup hierarchy among the 17 groups and ask to what extent the hierarchy is reflected in brain activity. Our data} provide a description of the visual system’s response to the complete set of symmetries in the two-dimensional plane. 
\textcolor{ForestGreen}{We do not} independently measure the response to \textit{P1}, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure \ref{fig:summary}), we can conclude that the relationships between P1 and all other groups, where \textit{P1} is the subgroup, are also preserved by the visual system. The subgroup relationships are in many cases not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. \textcolor{ForestGreen}{Previous brain-imaging studies have found evidence of parametric responses with the number of reflection symmetry folds \cite{keefe_2018, Sasaki_2005, makin_2016} and with the order of rotation symmetry \cite{RN1725}. Our study is the first demonstration that the brain encodes symmetry in this parametric fashion across every possible combination of different \textit{symmetry types}, and that this parametric encoding is also reflected in behavior.} Previous behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \citep{RN1253}, they tend to sort exemplars into fewer \textcolor{ForestGreen}{(4-12) sets than the number of wallpaper groups, often placing exemplars from different wallpaper groups in the same set \citep{RN172}. The two-interval forced choice approach we use in the current psychophysical experiment makes it possible to directly compare symmetry detection thresholds to the subgroup hierarchy, and reveals that not only can the 17 wallpaper groups be distinguished based on behavioral data, behavior largely follows the subgroup hierarchy.}

\textcolor{ForestGreen}{A large literature exists on the \textit{Sustained Posterior Negativity} (SPN), a characteristic negative-going waveform that is known to reflect responses to symmetry and other forms of regularity and structure \citep{makin_2016}. The SPN scales with the proportion of reflection symmetry in displays that contain a mixture of symmetry and noise \cite{makin_formation_2020, palumbo_scaling_2015}, and both reflection, rotation and translation can produce a measurable SPN \cite{makin_2013}. It has recently been demonstrated that a holographic model of regularity \citep{van_der_helm_1996}, can predict both SPN amplitude \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for dot patterns that contain symmetry and other types of regularity. The available evidence suggests that the SPN and our SSVEP measurements are two distinct methods for isolating the same symmetry-related brain response: When observed in the time-domain, the symmetry-selective odd-harmonic responses produce similarly sustained waveforms (see Figure \ref{fig:summary}), odd-harmonic SSVEP responses can be measured for dot patterns similar to those used to measure the SPN \citep{RN1954}, and the one event-related study on the wallpaper groups also found SPN-like waveforms \citep{RN1959}. Future work should more firmly establish the connection and determine if the SPN can capture similarly precise symmetry responses as the SSVEPs presented here. It would also be worthwhile to ask if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane.}

We observe a reliable correlation between our brain imaging and psychophysical data. This suggests that the two measurements reflect the same underlying symmetry representations in visual cortex. \textcolor{ForestGreen}{It should be noted that the correlation is relatively modest ($R^2 = 0.44$). This may be partly due to the fact that different individuals participated in the two experiments. It may also be related to the fact that participants where not doing a symmetry-related task during the SSVEP experiment, but instead monitored the stimuli for brief changes in contrast that occured twice per trial (see Methods). Previous brain imaging studies have found enhanced reflection symmetry responses when participants performed a symmetry-related task \citep{makin_formation_2020, Sasaki_2005, keefe_2018}. It is possible that adding a symmetry-related task to our SSVEP experiment would have produced measurements that reflected subgroup relationships to an even higher extent than what we observed. On the other hand, our results are already close to ceiling (see Figure \ref{fig:roc_style}) and adding a symmetry-related task may simply enhance SSVEP amplitudes overall without improving the discriminality of individual groups, as has been observed for reflection by \cite{keefe_2018}. Task-driven processing may be important for detecting symmetries that have been subject to perspective distortion, as suggested by SPN measurements \citep{makin_conditions_2015} and somewhat less clearly in a subsequent functional MRI study \citep{keefe_2018}. Future work in which behavioral and brain imaging data are collected from the same participants, and task is manipulated in the SSVEP experiment, will help further establish the connection between the two measurements, and elucidate the potential contribution of task-related top-down processing to the current results.}

@@

The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective \textcolor{ForestGreen}{(see above)} and occlusion. Near-regular textures, however, abound in the visual world, \textcolor{ForestGreen}{and can be modeled} as deformed versions of the wallpaper groups \citep{RN1519}. The correspondence between visual cortex responses and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \citep{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \citep{RN1768}. Field biologists have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \citep{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered \textit{supertextures}, exaggerated examples of the near-regular textures \textcolor{ForestGreen}{common in the natural world. If non-human animals employ a similar encoding strategy, they would be expected to be sensitive to symmetries in wallpaper groups. Recent functional MRI work in macaque monkeys offer some support for that: Macaque visual cortex responds parametrically to reflection and rotation symmetries in wallpaper groups, and the set of brain areas involved largely overlap those observed to be sensitive to symmetry in humans \citep{audurier_symmetry_2021}.} In human societies, visual artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \citep{RN1764}. Wallpaper groups are visually compelling, and symmetries have been widely used in human artistic expression going back to the Neolithic age \citep{RN1949}. If wallpapers are in fact supertextures, this prevalence may be a direct result of the strategy the human visual system has adopted for texture encoding.

\begin{figure}[tbp]
\centering
\includegraphics[width=0.5\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relationships that satisfy \textcolor{ForestGreen}{$p(\Delta >0 | data) > x$ for the SSVEP data and $p(\Delta <0 | data) > x$ for the psychophysical data}. We can see that if we take $x = 0.95$ as our threshold, the subgroup relationships are preserved in 56/63 = 89\% and \textcolor{ForestGreen}{48/63 = 76\%} of the comparisons for the primary SSVEP and threshold duration datasets, receptively. This compares to the \textcolor{ForestGreen}{32/63= 51\%} and \textcolor{ForestGreen}{22/63 = 35\%} for the SSVEP control datasets.}
\label{fig:roc_style}
\end{figure}

\subsection*{Participants}
Twenty-five participants (11 females, mean age 28.7 \textpm\ 3.3) took part in the EEG experiment. Their informed consent was obtained before the experiment under a protocol that was approved by the Institutional Review Board of Stanford University. 11 participants (8 females, mean age 20.73 \textpm\ 1.21) took part in the psychophysics experiment. All participants had normal or corrected-to-normal vision. Their informed consent was obtained before the experiment under a protocol that was approved by the University of Essex's Ethics Committee. \textcolor{ForestGreen}{There was no overlap in participants between the EEG and psychophysics experiments.}

@@

\subsection*{EEG Procedure}
Visual Evoked Potentials were measured using a steady-state design, in which \textit{P1} control images alternated with exemplar images from each of the 16 other wallpaper groups. Exemplar images were always preceded by their matched \textit{P1} control image. A single 0.83 Hz stimulus cycle consisted of a control \textit{P1} image followed by an exemplar image, each shown for 600 ms. A trial consisted of 10 such cycles (12 sec) over which 10 different exemplar images and matched controls from the same rotation group were presented. For each group type, the individual exemplar images were always shown in the same order within the trials. Participants initiated each trial with a button-press, which allowed them to take breaks between trials. Trials from a single wallpaper group were presented in blocks of four repetitions, which were themselves repeated twice per session, and shown in random order within each session. To control fixation, the participants were instructed to fixate a small white cross in the center of display. To control vigilance, a contrast dimming task was employed. Two times per trial, an image pair \textcolor{ForestGreen}{(control \textit{P1} plus exemplar)} was shown at reduced contrast. \textcolor{ForestGreen}{P}articipants were instructed to press a button on a response pad \textcolor{ForestGreen}{whenever they noticed a contrast change. Reaction times were not taken into account and participants were told to respond at their own pace while being as accurate as possible}. We adjusted the \textcolor{ForestGreen}{reduction in contrast} such that average accuracy for each participant was kept at ~85\% correct, in order to keep the difficulty of the vigilance \textcolor{ForestGreen}{task} at a constant level.

>>>>

\begin{abstract}
Symmetries are present at many scales in natural scenes. Humans and other animals are highly sensitive to visual symmetry, and symmetry contributes to numerous domains of visual perception. The four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined into exactly 17 distinct regular textures. These \textit{wallpaper groups} represent the complete set of symmetries in 2D images. The current study seeks to provide a more comprehensive description of responses to symmetry in the human visual system, by collecting both brain imaging (Steady-State Visual Evoked Potentials measured using high-density EEG) and behavioral (symmetry detection thresholds) data using the entire set of wallpaper groups. This allows us to probe the hierarchy of complexity among wallpaper groups, in which simpler groups are subgroups of more complex ones. We find that both behavior and brain activity preserve the hierarchy almost perfectly: Subgroups consistently produce lower amplitude symmetry-specific responses in visual cortex and require longer presentation durations to be reliably detected. These findings expand our understanding of symmetry perception by showing that the human brain encodes symmetries with a high level of precision and detail. This opens new avenues for research on how fine-grained representations of regular textures contribute to natural vision.
\end{abstract}

Symmetries are abundant in natural and man-made environments, due to a complex interplay of physical forces that govern pattern formation in nature. Sensitivity to symmetry has been demonstrated in a number of species, includes bees \citep{giurfa_insect_symmetry_1996}, fish \citep{morris_swordtail_symmetry_1998,schluter_sailfin_symmetry_1998}, birds \citep{RN1337,swaddle_symmetry_finches_1994} and dolphins \citep{von_fersen_dolphin_symmetry_1992}, and may be used as a cue for mate selection in many species \citep{swaddle_asymmetry_review_1999} including humans \citep{rhodes_facial_1998}. Humans cultures have created and appreciated symmetrical patterns throughout history, and since the gestalt movement of the early 20th century, symmetry has been recognized as important for visual perception. Symmetry contributes to the perception of shapes \citep{RN1311,RN1682}, scenes \citep{apthorp_symmetry_numerosity} and surface properties \citep{RN1166}. This literature is almost exclusively based on stimuli in which one or more symmetry axes are placed at a single point in the image. Focus has been on mirror symmetry or \textit{reflection}, with relatively few studies including the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection} \citep{wagemans_1998} - perhaps because reflection has been found to be more perceptually salient \citep{mach_1959,royer_1981,palmer_1991,ogden_2016,hamada_1988} and produce more brain activity \citep{makin_2013,makin_2014,makin_2012,wright_2015}. In the current study, we take a different approach by investigating visual processing of regular textures in which combinations of the four fundamental symmetries tile the 2D plane.

In the two spatial dimensions relevant for images, symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \citep{RN1562,RN1563,RN1425}. Previous work on a subset of four of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \citep{RN1725}. This effect was also robust when symmetry responses were measured with electroencephalography (EEG) using both Steady-State Visual Evoked Potentials (SSVEPs)\citep{RN1725} and Event-Related Potentials \citep{RN1959}. The SSVEP technique uses periodic visual stimulation to produce a periodic brain response that is confined to integer multiples of the stimulation frequency known as harmonics. SSVEP response harmonics can be isolated in the frequency domain and depending on the specific design, different harmonics will express different aspects of the brain response. \citep{RN1540}. Here we extend on the previous work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \citep{RN1425}. The wallpaper groups differ in the number and kind of these transformations and we can uniquely refer to different groups using crystallographic notation. In brief, most groups are notated by $PXZ$, where $X\in\{1,2,3,4,6\}$ indicates the highest order of rotation symmetry and $Z\in\{m, g\}$ indicates whether the pattern contains reflection (m) or glide reflection (g). For example, $P4$ contains rotation of order 4, while $P4MM$ contains rotation of order 4 and two reflections. By convention, many of the groups are given shortened names: for example, $P4MM$ is usually referred to as $P4M$, as the second reflection can be deduced from the presence of rotation of order 4 alongside a reflection. Two of the groups start with a $C$ rather than a $P$, ($CM$ and $CMM$) which indicates that the symmetries are specified relative to a cell that itself contains repetition. Full details of the naming convention can be found on \href{https://en.wikipedia.org/wiki/Wallpaper_group}{wikipedia} and examples of the wallpaper groups are shown in Figures \ref{fig:subgroups} and \ref{fig:summary}.

In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \citep{RN1425}. The full list of subgroup relationships is listed in Section 1.4.2 of the Supplementary Material. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the supergroup \citep{RN1425}. As an example, let us consider groups \textit{P2} and \textit{P6} (see Figure \ref{fig:subgroups}B). If we ignore the translations in two directions that both groups share, group \textit{P6} consists of the set of rotations \{$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$\}, in which \textit{P2} \{$0^{\circ}$, $180^{\circ}$\} is contained. \textit{P2} is thus a subgroup of \textit{P6}, and \textit{P6} can be generated by combining \textit{P2} with rotations \{$0^{\circ}$, $120^{\circ}$, $240^{\circ}$\}. Because \textit{P2} is repeated three times in \textit{P6}, \textit{P2} is a subgroup of \textit{P6} with index 3 \citep{RN1425}. Similarly, \textit{PMM} contains two reflections and rotations \{$0^{\circ}$, $180^{\circ}$\}. \textit{PMM} can be generated by adding an additional reflection to both \textit{P2} (\{$0^{\circ}$, $180^{\circ}$\}) and \textit{PM} (one reflection), so \textit{P2} and \textit{PM} are both subgroups of \textit{PMM} with index 2 (see Figure \ref{fig:subgroups}C). The 17 wallpaper groups thus obey a hierarchy of complexity where simpler groups are subgroups of more complex ones \citep{RN1711}.

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure1.pdf}
\caption{Subgroup relationships with indices 2 (solid lines) and 3 (dashed line) are shown in (A). All other relationships can be inferred by identifying the shortest path through the hierarchy, and multiplying the subgroup indices. For example, \textit{P1} is related to \textit{P6} through \textit{P6}\textrightarrow\textit{P3} (index 2) and \textit{P3}\textrightarrow\textit{P1} (index 3) so \textit{P1} is also a subgroup of \textit{P6} with index $3 \times 2 = 6$. We also show enlarged versions of some of the subgroup relationships involving \textit{P6} (B, shown in red) and \textit{PMM} (C, shown in blue) and highlight the symmetries within the subgroups to emphasize how the supergroup can be generated by adding additional transformations to the subgroup.}
\label{fig:subgroups}
\end{figure}

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \citep{Sasaki_2005, Tyler_2005} and rotation \citep{RN1725}, the visual system represents the hierarchical structure of the 17 wallpaper groups, and thus every combination of the four fundamental symmetries (rotation, reflection, translation, glide reflection) which comprise the set of regular textures. Both SSVEP amplitudes and symmetry detection thresholds preserve the hierarchy of complexity among the wallpaper groups that is captured by the subgroup relationships \citep{RN1711}. For the SSVEP, this remarkable consistency was specific to the odd harmonics of the stimulus frequency that are known to capture the symmetry-specific response \citep{RN1725} and to electrodes in a region-of-interest (ROI) over occipital cortex. When the same analysis was done using the odd harmonics from electrodes over parietal cortex (Supplementary Figure 3.3) or even harmonics from electrodes over occipital cortex (Supplementary Figure 3.4), the data was substantially less consistent with the subgroup relationships (yellow and green lines, Figure 5).

The current study uses 16 distinct wallpaper groups, while previous neuroimaging studies focused on a subset of 4 \citep{RN1725,RN1959}. This represents a significant conceptual advance, because it makes it possible to investigate the complete subgroup hierarchy among the 17 groups and ask to what extent the hierarchy is reflected in brain activity. Our data provide a description of the visual system’s response to the complete set of symmetries in the two-dimensional plane. We do not independently measure the response to \textit{P1}, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure \ref{fig:summary}), we can conclude that the relationships between P1 and all other groups, where \textit{P1} is the subgroup, are also preserved by the visual system. The subgroup relationships are in many cases not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. Previous brain-imaging studies have found evidence of parametric responses with the number of reflection symmetry folds \cite{keefe_2018, Sasaki_2005, makin_2016} and with the order of rotation symmetry \cite{RN1725}. Our study is the first demonstration that the brain encodes symmetry in this parametric fashion across every possible combination of different \textit{symmetry types}, and that this parametric encoding is also reflected in behavior. Previous behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \citep{RN1253}, they tend to sort exemplars into fewer (4-12) sets than the number of wallpaper groups, often placing exemplars from different wallpaper groups in the same set \citep{RN172}. The two-interval forced choice approach we use in the current psychophysical experiment makes it possible to directly compare symmetry detection thresholds to the subgroup hierarchy, and reveals that not only can the 17 wallpaper groups be distinguished based on behavioral data, behavior largely follows the subgroup hierarchy.

A large literature exists on the \textit{Sustained Posterior Negativity} (SPN), a characteristic negative-going waveform that is known to reflect responses to symmetry and other forms of regularity and structure \citep{makin_2016}. The SPN scales with the proportion of reflection symmetry in displays that contain a mixture of symmetry and noise \cite{makin_formation_2020, palumbo_scaling_2015}, and both reflection, rotation and translation can produce a measurable SPN \cite{makin_2013}. It has recently been demonstrated that a holographic model of regularity \citep{van_der_helm_1996}, can predict both SPN amplitude \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for dot patterns that contain symmetry and other types of regularity. The available evidence suggests that the SPN and our SSVEP measurements are two distinct methods for isolating the same symmetry-related brain response: When observed in the time-domain, the symmetry-selective odd-harmonic responses produce similarly sustained waveforms (see Figure \ref{fig:summary}), odd-harmonic SSVEP responses can be measured for dot patterns similar to those used to measure the SPN \citep{RN1954}, and the one event-related study on the wallpaper groups also found SPN-like waveforms \citep{RN1959}. Future work should more firmly establish the connection and determine if the SPN can capture similarly precise symmetry responses as the SSVEPs presented here. It would also be worthwhile to ask if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane.

We observe a reliable correlation between our brain imaging and psychophysical data. This suggests that the two measurements reflect the same underlying symmetry representations in visual cortex. It should be noted that the correlation is relatively modest ($R^2 = 0.44$). This may be partly due to the fact that different individuals participated in the two experiments. It may also be related to the fact that participants where not doing a symmetry-related task during the SSVEP experiment, but instead monitored the stimuli for brief changes in contrast that occured twice per trial (see Methods). Previous brain imaging studies have found enhanced reflection symmetry responses when participants performed a symmetry-related task \citep{makin_formation_2020, Sasaki_2005, keefe_2018}. It is possible that adding a symmetry-related task to our SSVEP experiment would have produced measurements that reflected subgroup relationships to an even higher extent than what we observed. On the other hand, our results are already close to ceiling (see Figure \ref{fig:roc_style}) and adding a symmetry-related task may simply enhance SSVEP amplitudes overall without improving the discriminality of individual groups, as has been observed for reflection by \cite{keefe_2018}. Task-driven processing may be important for detecting symmetries that have been subject to perspective distortion, as suggested by SPN measurements \citep{makin_conditions_2015} and somewhat less clearly in a subsequent functional MRI study \citep{keefe_2018}. Future work in which behavioral and brain imaging data are collected from the same participants, and task is manipulated in the SSVEP experiment, will help further establish the connection between the two measurements, and elucidate the potential contribution of task-related top-down processing to the current results.

@@

The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective (see above) and occlusion. Near-regular textures, however, abound in the visual world, and can be modeled as deformed versions of the wallpaper groups \citep{RN1519}. The correspondence between visual cortex responses and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \citep{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \citep{RN1768}. Field biologists have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \citep{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered \textit{supertextures}, exaggerated examples of the near-regular textures common in the natural world. If non-human animals employ a similar encoding strategy, they would be expected to be sensitive to symmetries in wallpaper groups. Recent functional MRI work in macaque monkeys offer some support for that: Macaque visual cortex responds parametrically to reflection and rotation symmetries in wallpaper groups, and the set of brain areas involved largely overlap those observed to be sensitive to symmetry in humans \citep{audurier_symmetry_2021}. In human societies, visual artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \citep{RN1764}. Wallpaper groups are visually compelling, and symmetries have been widely used in human artistic expression going back to the Neolithic age \citep{RN1949}. If wallpapers are in fact supertextures, this prevalence may be a direct result of the strategy the human visual system has adopted for texture encoding.

\begin{figure}[tbp]
\centering
\includegraphics[width=0.5\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relationships that satisfy $p(\Delta >0 | data) > x$ for the SSVEP data and $p(\Delta <0 | data) > x$ for the psychophysical data. We can see that if we take $x = 0.95$ as our threshold, the subgroup relationships are preserved in 56/63 = 89\% and 48/63 = 76\% of the comparisons for the primary SSVEP and threshold duration datasets, receptively. This compares to the 32/63= 51\% and 22/63 = 35\% for the SSVEP control datasets.}
\label{fig:roc_style}
\end{figure}

\subsection*{Participants}
Twenty-five participants (11 females, mean age 28.7 \textpm\ 3.3) took part in the EEG experiment. Their informed consent was obtained before the experiment under a protocol that was approved by the Institutional Review Board of Stanford University. 11 participants (8 females, mean age 20.73 \textpm\ 1.21) took part in the psychophysics experiment. All participants had normal or corrected-to-normal vision. Their informed consent was obtained before the experiment under a protocol that was approved by the University of Essex's Ethics Committee. There was no overlap in participants between the EEG and psychophysics experiments.

@@

\subsection*{EEG Procedure}
Visual Evoked Potentials were measured using a steady-state design, in which \textit{P1} control images alternated with exemplar images from each of the 16 other wallpaper groups. Exemplar images were always preceded by their matched \textit{P1} control image. A single 0.83 Hz stimulus cycle consisted of a control \textit{P1} image followed by an exemplar image, each shown for 600 ms. A trial consisted of 10 such cycles (12 sec) over which 10 different exemplar images and matched controls from the same rotation group were presented. For each group type, the individual exemplar images were always shown in the same order within the trials. Participants initiated each trial with a button-press, which allowed them to take breaks between trials. Trials from a single wallpaper group were presented in blocks of four repetitions, which were themselves repeated twice per session, and shown in random order within each session. To control fixation, the participants were instructed to fixate a small white cross in the center of display. To control vigilance, a contrast dimming task was employed. Two times per trial, an image pair (control \textit{P1} plus exemplar) was shown at reduced contrast. Participants were instructed to press a button on a response pad whenever they noticed a contrast change. Reaction times were not taken into account and participants were told to respond at their own pace while being as accurate as possible. We adjusted the reduction in contrast such that average accuracy for each participant was kept at ~85\% correct, in order to keep the difficulty of the vigilance task at a constant level.
